Tools for Identifying Partisan Gerrymandering (PA)

### 2019-06-06 Latex/_maintext.tex
Update _maintext.tex
Latex files update

    In addition to the rejected 2011 and the court-ordered 2018 map, we also address the partisan and other features of two maps offered to the Pennsylvania Court as a proposed remedy for constitutional violation found in the 2011 map: the \textit{Joint Plan} offered by Republican legislators from both chambers, and the congressional map proposed by the Democratic governor (\textit{Gov. Wolf}). Here we focus on the question of whether either of these plans was a \textit{stealth gerrymander} done to aid the party drawing the map.

@@

    As noted above, to contrast the Court-ordered Remedial plan with the 2011 Enacted plan, and to make further comparisons to the two most politically important remedy plans offered to the Court, the Joint Plan proposed by the Republicans leaders of the two legislative chambers, and the Governor’s proposed plan as a Democratic alternative, we will look at a number of features of each of these four plans as a whole.  We do not, however, discuss district-specific criteria for identifying gerrymandering in these maps, since, although some district-specific evidence was presented about the 2011 map in the trial testimony by an expert who was highly knowledgeable about communities of interest in the state, the Pennsylvania Supreme Court opinion itself offers a \underline{statewide} standard for a finding of unconstitutional partisan gerrymandering. 

@@

% County Splits, Compactness Scores of the Plans

@@

% -- END TABLE -- END TABLE -- END TABLE -- END TABLE -- END TABL %

@@

        \footnote{We report the Republican share of the two-party vote ($\delta_{i}$). Districting plans are represented by $\mathcal D $, (e.g., $\mathcal D_{enacted}$, $\mathcal D_{remedial}$, \dots, $\mathcal D_{j}$), and each election in year $ y $ has a district vote distribution $ \Delta_{y} = [\delta_{y1}, \delta_{y2}, \dots, \delta_{yi}] $. To find the overall state-wide vote, we calculate the average district vote share, $ \bar{\Delta}_{\mathcal D y} = \frac{1}{n}\sum\limits_{i=1}^{n} \delta_{yi} $. By averaging, we reduce the influence of turnout variation among districts. This district-specific average is a useful state-wide estimate of voter sentiment though it is in part effected by the particular configuration of districts being examined \citep[see e.g.][cf. \cite{Grofman_et_al_1997__IntedgratedPerspective_ES}]{Kastellec_et_al_2008_PS}.In particular, incumbency advantage effects affect our ability to accurately predict vote outcomes. In ancillary work not reported here, we have looked at changes in congressional incumbency advantage in Pennsylvania over multiple decades.}

@@

    Our confidence in the five election projection method is further buttressed because, when we applied this method to predict the results of the 2018 election under the court-ordered remedial map. Although the method had been developed prior to the 2018 election, using data no more recent than 2016, and not including any incumbency advantage, it predicted that the court-ordered remedial plan would have produced an outcome of between 5R-12D and 11R-7D.

@@

        \footnote{After uniformly shifting each district's composite such that the statewide average equalled the actual 2018 statewide average. The large range is indicative of the number of competitive districts that Republican candidates won. Had we modeled incumbency advantage instead of the `normal vote', the two Republican incumbents who won by less than 2 percentage points, along with the one who won by less than 3 percentage points would have made the actual 9R-9D outcome the likely composite outcome.}

@@

% =================================================================
% -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE -- TABLE  % 
% -----------------------------------------------------------------
% Measures of Gerrymandering for the Four Considered Plans
% ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••• 
%    
\begin{center} INSERT TABLE \ref{tab:gerry} ABOUT HERE \end{center}

@@

% Probabilistic Projections of Partisan Outcomes for Four Plans

@@

\begin{center} INSERT TABLE \ref{tab:prob} ABOUT HERE \end{center}

@@

    The findings from Table \ref{tab:gerry} are very clear! Regardless of which metric we examine, the 2011 map is the most gerrymandered. Not surprisingly, the nature of the bias is in a pro-Republican direction. Given the cumulative weight of all the evidence, the 2011 congressional map was clearly an egregious pro-Republican gerrymander. In contrast, as we might expect, the court-ordered plan, prepared by a respected expert whose instructions were to satisfy good government criteria, was by a considerable margin the closest to a perfectly neutral plan in partisan terms according to all four measures. 

@@

    Finally, we turn to the plan offered by the Democratic governor, Governor Wolf. To our surprise, this plan turned out also to have a pro-Republican tilt, nearly to the same extent as the Joint Legislative plan for three of the four metrics in Table \ref{tab:gerry}. This puzzling finding suggests that the Governor might have been under pressure from Democratic incumbents not to reduce their expected vote margins, or that he sincerely intended it to be a compromise that the Republican legislature might agree to. However, there is one important difference between the Joint plan and that of the Governor: among the seven districts likely to be won by Democrats, the Democratic percentages are cut more narrowly in the Republican plan than in the Governor’s plan in three of the districts so that, in a good Republican year, Republicans will do better under their plan than under the Democratic plan.

@@

    For example, in the first set of 500 simulations in his expert witness report, based only on traditional redistricting criteria, but not including voting rights protections, Professor Chen found that none resulted in the 13R-5D split as happened in 2012 and subsequent years. The mode was 9R-9D.  

@@

    Recognizing the potential for so-called ``natural gerrymandering", we wish to highlight a potentially important distinction between \textit{neutral} plans and \textit{fair} plans -- each of which reflects a different approach to defining the baseline for what constitutes gerrymandering. Neutral plans refer to those that are drawn entirely with respect to traditional good government criteria, with no attention paid to partisan considerations. One way to define partisan gerrymandering is with respect to a baseline defined by the set of feasible neutral plans -- and this was how one of the plaintiff’s experts in the Pennsylvania case, Professor Jowei Chen \citep[cf.][]{Chen2013} shaped his testimony in the Pennsylvania case, and in other cases in which he has testified.

@@

    In contrast, we may wish to use the term ``fair" plans as ones that seek to compensate for the difference in level of partisan concentration. Fair plans do not seek to achieve proportional representation. Rather, they seek to achieve partisan symmetry. A fair plan would yield a proportional result when the vote shares are evenly split 50-50, but at points beyond 50-50, such as when party R wins 70\% of the districts with only 55\% of the vote, party L can similarly win 70\% of the districts if it had received 55\% of the vote. 

@@

    But the Pennsylvania Supreme Court opted not to do that, but to instead preserve county lines and draw a good government map. We can therefore identify the Court remedial plan as a ``neutral" plan.

@@

    While the Pennsylvania court opinion is limited to Pennsylvania, and thus it might have seemed of importance only in Pennsylvania, in footnote 71 of the Opinion (slip op. pp.116-117), the court took what we regard as a rather unusual step. It issued what can only be called an invitation to other state courts to use the same logic it used to invalidate partisan gerrymanders in their own state. As we noted earlier, the Court pointed out that there are twelve states whose constitutions contain election clauses identical to the Pennsylvania charter, requiring elections to be “free and equal”: Arizona, Arkansas, Delaware, Illinois, Indiana, Kentucky, Oklahoma, South Dakota, Oregon, Tennessee, Washington, and Wyoming.

@@

    However, only a handful of these states are ripe for partisan gerrymandering challenges -- in that (ca. 2017) three lack unified party control of the state, two are single district states, and some already have a commission drawing plans and, in others, \textit{indicia} of gerrymandering are missing.

@@

    In looking to the future, we should also note that some states that are generally regarded as among the most pernicious partisan gerrymanders, Michigan, Ohio, North Carolina, Wisconsin are not included among the twelve, nor is Maryland. On the other hand, we should also note that a “free and equal” elections clause is not the only avenue state courts might use to attack partisan gerrymanders in the future. As University of Kentucky College of Law Professor Joshua Douglas has pointed out, virtually every state constitution protects voting rights more explicitly than the U.S. Constitution does \citep{Douglas2014_RightToVote}. In addition to the thirteen states that require elections to be ``free and equal”, an additional thirteen have state constitutional provisions that require elections to be ``free" or ``free and open”, and this clause could, in principle, be used in exactly the same way as the ``free and equal” clause. 

@@

    Such a clause is found in a number of the states widely regarded as having the worst congressional gerrymandering. 

@@

    Accordingly, some scholars have suggested turning to turn away from the federal judiciary altogether and instead focusing on state-level litigation \citep{Wang_et_al_2019_Labortories_UPJCL}. We believe that the \textit{LWV} decision and the further comparative analyses of alternative plans presented in the previous section of this paper will help other state courts navigate their way to decisions that strike down some partisan gerrymanders as unconstitutional, while allowing others to remain in place on the grounds that either they are not that severe, or they are unlikely to be lasting in that they have sufficiently many competitive seats as to be responsive to realistic changes in voting patterns. Now that litigation has been successful in state courts on the grounds of egregious deviations from good government districting criteria, it seems plausible that in the next wave of redistricting, mapmakers may create covert gerrymanders like we've described in this essay. If so, there remains the issue of whether state courts have the legal tools to deal with these ``stealth gerrymander”.

@@

    According to all four measures, the Joint Legislative plan offered by Republican leaders is nearly as much a gerrymandered map -- in a pro-Republican direction -- as is the 2011 map. As noted above, we believe it appropriate to label this plan as a stealth gerrymander since it achieves partisan purposes while (mostly) providing the visual appearance of a good government map (see Figure \ref{fig:maps}). However, having labeled the Joint Legislative plan a \textit{stealth gerrymander}, it is worth reminding readers of two important points. 
    First, had this plan been the original plan adopted by Republicans, it is very much an open question as to whether it would have been rejected by the Pennsylvania Supreme Court. Even though, in partisan terms, it is nearly identical to the 2011 plan, it is considerably more consistent with good government criteria (see Table \ref{tab:summaries}) -- not that much worse than the Court Remedial plan -- that it would have required detailed political election analyses like those we have done here to demonstrate its stealth gerrymander features, plus a court willing to rule on political grounds rather than good government grounds that it was unconstitutional. Because the Joint Legislative plan was not passed by the State of Pennsylvania legislature, the Court felt no need for particular deference to it (or to the Governor's plan, for that matter). Thus, the Court was not bound by the normal supposition with respect to districting that a plan authorized by the state need not be the ``best possible," but only unconstitutional. That ``deference" to legislative judgments and the criteria they reflect, might well have tipped the balance toward acceptance of the Joint Legislative plan were it offered as having been fully sanctioned by the State of Pennsylvania (the duly elected legislature and Governor). 

@@

@@

	Thus we have some optimism that state courts can provide a venue to check partisan gerrymandering, even when it comes in the form of a \textit{stealth gerrymander}. But we have also placed a number of caveats on any claim that state courts can compensate fully for any continued failure of the federal courts to act. Citizen-based reforms such as referendum and initiatives may also help with reducing egregious gerrymandering in states where the passage of such measures is feasible. Ultimately, there is no substitute for a federal standard that guarantees each citizen in the United States \textit{equal protection under the law}. Absent the United States Supreme Court setting this standard, state courts offer one promising pathway towards ending undemocratic partisan gerrymandering.

@@

% %        ================================================================= 
% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % ------------------------------------------------------------------------
% 		\footnote{A long literature in political science has evaluated and confirmed the presence and increased incumbency advantage \citep{Abramowitz1991,Levitt1997,Ansolabehere2002b,Gelman_Huang_2008}. Recent scholarship, however, has shown a decrease in the advantage towards zero \citep{Jacobson2015_JOP}.}
% % ------------------------------------------------------------------------ 
% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% %        =================================================================
%     A well-financed incumbent tends to suppress quality competition \citep{Jacobson2015_JOP}, vulnerable incumbents who can sense the political winds opt against running for re-election \citep{Jacobson_Kernell_1981,Ansolabehere_Snyder_2004_LSQ}, and electoral tides and varying turnout can distort the state-wide average. Midterm elections often have vastly different electorates than presidential elections \citep{Abramowitz_et_al_1986_AJPS}, while the midterm pattern of ``surge and decline" (what goes up must come down principle) reduces the sitting Presidents' party's vote share \citep{CampbellA1960}. 
%  \par
% 	What we are after is a measure of the ``normal vote''. The normal vote is the underlying partisanship of a district \citep{Converse1966}. It is sort of a mystical number since it is uncovered and unobserved, but powerful theoretically. Map builders implicitly assume a normal vote when constructing a district plan. We believe that state-wide elections offer the most useful approximation of the normal vote.
% %        ================================================================= 
% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % ------------------------------------------------------------------------
% 		\footnote{We particularly believe this is true when the votes between elections are highly correlated, and the district-level variation is low, as it was in Pennsylvania from 2012 to 2016. The more stable the electorate is, the more confidence we have that state-wide elections are revealing the normal vote \citep{Ansolabehere_et_al_2000_AJPS}.}
% % ------------------------------------------------------------------------ 
% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% %        =================================================================
% 	 All districts vote on the same candidates; hence outcome affects all voters equally. The effects of incumbency can distort the vote in congressional races in a way that doesn't resemble the actual partisan preferences at the district-level. State-wide elections eliminate the need to impute or otherwise deal with uncontested districts. Taking an average of the state-wide elections provides additional reason to believe it's capturing a normal vote due to a regression to the mean effects. Justices in \textit{Vieth} offered skepticism in using elections that have contrasting results; we do not share this view, instead seeing it as an important source of variation that helps to recognize the underlying partisanship in which \textit{should} be used in measuring bias in districting plans. Additionally, averaging over multiple races and years helps to reduce the bias imposed by candidate effects.
% \par
% 	We construct the five-election composite of state-wide elections by aggregating voting district level (precincts) data to the congressional district level.
% %        ================================================================= 
% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % ------------------------------------------------------------------------
% 		\footnote{We wrote a function in R to assign all VTDs to a district based on the centroid of the VTD. This method creates a slight difference between it and the actual vote, but averages just -0.06\% per district. Most of the mis-allocation is between two adjacent districts, District 1 and District 13, both overwhelmingly Democratic districts.}
% % ------------------------------------------------------------------------ 
% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% %        =================================================================
% 	Encouragingly, our five-election composite results in the same 13R-5D result that the 2016 congressional races delivered.
% %        ================================================================= 
% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % ------------------------------------------------------------------------
% 		\footnote{Whether we take the sum of the five elections or average the five elections, both methods deliver the same relative distributions.}
% % ------------------------------------------------------------------------ 
% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% %        =================================================================

@@

% %        ================================================================= 
% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % ------------------------------------------------------------------------	
% 	We calculate this average only for districts contested in 2016, since uncontested results have significant deviations. Including the uncontested districts increases this mean difference to just 3.93\%. We believe our composite measure offers better representation of the true partisanship in these districts. Individually, only the Auditor's race correctly predicts all the districts correctly, but by either summing or by taking the average of the five races, both methods correctly predict each district's outcome.
% % ------------------------------------------------------------------------ 
% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% %        =================================================================
%     We imputed the values 25\% and 75\% for uncontested or essentially uncontested races. Rather than taking the projected outcome we arrive at from this method as a single value, we looked at the sensitivity of this projection to the empirical range of inter-election swing, simulating outcomes 1,000 times. To arrive at our estimate for inter-election variation that can't be attributed to the normal vote or incumbency advantage, we estimate the equation using congressional elections in Pennsylvania in the 2004-16 time period:
%         \begin{align*}
%         \mathrm{Vote}_{ty} = \alpha + \beta_{1} \mathrm{Vote}_{yt-1} + \beta_{2} \mathrm{Incumbency}_{yt} + \beta_{4} \mathrm{Uncontestedness}_y{t} + \epsilon
%         \end{align*}
%     We are interested in $\epsilon$, which is the unexplained variance, which includes inter-election swing. We run this equation separately for each of the non-redistricting years and then average the coefficients. We estimate incumbency advantage to average 4.27\% for this time period, while inter-election variance is 4.46\%. When simulating elections, we use the composite as the normal vote, but add in for each district a random draw from a normal distribution centered at the district vote with a standard deviation of $\epsilon$. We can then calculate statistics for each of these simulated elections to find a distribution for each measure of gerrymandering.
% \par
%     Table \ref{tab:gerry} reports the same four measures of partisan effects found in Table \ref{tab:congsum} but now for all four maps, using the projected statewide vote from our five-election composite, under the methodology described above. We also provide a measure of variability to show what range of reasonable outcomes for each of the plans and identify when gerrymandering effects are persistent.

% \par
%     As we might also expect, again according to all four measures, the next most gerrymandered map -- also in a pro-Republican direction -- is the \textit{Joint Legislative} plan offered by Republican legislative leaders as a remedy plan. The partisan gerrymandering indicators for this plan are sufficiently close to those of the 2011 map, and sufficiently far from those of the court-ordered map, that we believe it appropriate to label this plan as a ``\textit{stealth gerrymander}" since it achieves partisan purposes while (mostly) providing the visual appearance of a good government map (see \ref{fig:maps}). 
% \par
%     However, having labeled the Joint Legislative plan a \textit{stealth gerrymander}, it is worth reminding readers that, had this plan been the original plan adopted by Republicans, it is very much an open question as to whether it would have been rejected by the Pennsylvania Supreme Court. Even though, in partisan terms, it is nearly identical to the 2011 plan, it is considerably more consistent with good government criteria (see Table \ref{tab:summaries}) -- not that much worse than the Court Remedial plan -- that it would have required detailed political election analyses like those we have done here to demonstrate its \textit{stealth gerrymander} features, plus a court willing to rule specifically on grounds that it was a \textit{partisan gerrymander}.
% \par
%     Because the Joint Legislative plan was not passed by the State of Pennsylvania legislature, the Court felt no need for particular deference to it (or to the Governor’s plan, for that matter). Thus, the Court was not bound by the normal supposition with respect to districting that a plan authorized by the state need not be the ``best possible," but only \underline{NOT} unconstitutional. That ``deference" to legislative judgments and the criteria they reflect, might well have tipped the balance toward acceptance of the Joint Legislative plan were it offered as having been fully sanctioned by the State of Pennsylvania (the duly elected legislature and Governor).

>>>>

    In addition to the rejected 2011 and the court-ordered 2018 map, we also address the partisan and other features of two maps offered to the Pennsylvania Court as a proposed remedy for constitutional violation found in the 2011 map: the \textit{joint legislative} plan offered by Republican legislators from both chambers, and the congressional map proposed by the Democratic governor (\textit{Gov. Wolf}). Here we focus on the question of whether either of these plans was a \textit{stealth gerrymander} done to aid the party drawing the map.

@@

    As noted above, to contrast the Court-ordered Remedial plan with the 2011 Enacted plan, and to make further comparisons to the two most politically important remedy plans offered to the Court, the joint plan proposed by the Republicans leaders of the two legislative chambers, and the Governor’s proposed plan as a Democratic alternative, we will look at a number of features of each of these four plans as a whole.  We do not, however, discuss district-specific criteria for identifying gerrymandering in these maps, since, although some district-specific evidence was presented about the 2011 map in the trial testimony by an expert who was highly knowledgeable about communities of interest in the state, the Pennsylvania Supreme Court opinion itself offers a \underline{statewide} standard for a finding of unconstitutional partisan gerrymandering. 

@@

\input{Tables/_tab_summaries.tex}

@@

@@

        \footnote{We report the Republican share of the two-party vote ($\delta_{i}$). Districting plans are represented by $\mathcal D $, (e.g., $\mathcal D_{enacted}$, $\mathcal D_{remedial}$, \dots, $\mathcal D_{j}$), and each election in year $ y $ has a district vote distribution $ \Delta_{y} = [\delta_{y1}, \delta_{y2}, \dots, \delta_{yi}] $. To find the overall state-wide vote, we calculate the average district vote share, $ \bar{\Delta}_{\mathcal D y} = \frac{1}{n}\sum\limits_{i=1}^{n} \delta_{yi} $. By averaging, we reduce the influence of turnout variation among districts. This district-specific average is a useful state-wide estimate of voter sentiment though it is in part effected by the particular configuration of districts being examined \citep[see e.g.][cf. \cite{Grofman_et_al_1997__IntedgratedPerspective_ES}]{Kastellec_et_al_2008_PS}. In particular, incumbency advantage effects affect our ability to accurately predict vote outcomes. In ancillary work not reported here, we have looked at changes in congressional incumbency advantage in Pennsylvania over multiple decades.}

@@

    Our confidence in the five election projection method is further buttressed because, when we applied this method to predict the results of the 2018 election under the court-ordered remedial map. Although the method had been developed prior to the 2018 election, using data no more recent than 2016, and not including any incumbency advantage, it predicted that the court-ordered remedial plan would have produced an outcome of 7R-11D.

@@

        \footnote{After uniformly shifting each district's composite such that the statewide average equalled the actual 2018 statewide average. The large range is indicative of the number of competitive districts that Republican candidates won. Had we modeled incumbency advantage instead of the `normal vote', the two Republican incumbents who won re-election by less than 2 percentage points, along with the one who won by less than 3 percentage points would have made the actual 9R-9D outcome the likely predicted outcome.}

@@

@@

\input{Tables/_tab_gerry.tex}
\begin{center}\textbf{INSERT TABLE \ref{tab:gerry} ABOUT HERE} \end{center}

@@

\input{Tables/_tab_prob.tex}
\begin{center}\textbf{INSERT TABLE \ref{tab:prob} ABOUT HERE} \end{center}

@@

    The findings from Table \ref{tab:gerry} are very clear; Regardless of which metric we examine, the 2011 map is the most gerrymandered. Not surprisingly, the nature of the bias is in a pro-Republican direction. Given the cumulative weight of all the evidence, the 2011 congressional map was clearly an egregious pro-Republican gerrymander. In contrast, as we might expect, the court-ordered plan, prepared by a respected expert whose instructions were to satisfy good government criteria, was by a considerable margin the closest to a perfectly neutral plan in partisan terms according to all four measures. 

@@

    Finally, we turn to the plan offered by the Democratic governor, Governor Wolf. To our surprise, this plan turned out also to have a pro-Republican tilt, nearly to the same extent as the joint legislative plan for three of the four metrics in Table \ref{tab:gerry}. This puzzling finding suggests that the Governor might have been under pressure from Democratic incumbents not to reduce their expected vote margins, or that he sincerely intended it to be a compromise that the Republican legislature might agree to. However, there is one important difference between the joint plan and that of the Governor: among the seven districts likely to be won by Democrats, the Democratic percentages are cut more narrowly in the Republican plan than in the Governor’s plan in three of the districts so that, in a good Republican year, Republicans will do better under their plan than under the Democratic plan.

@@

    For example, in the first set of 500 map simulations in his expert witness report, based only on traditional redistricting criteria, but not including voting rights protections, Professor Chen found that none resulted in the 13R-5D split as happened in 2012 and subsequent years. The mode was 9R-9D.

@@

    Recognizing the potential for so-called ``natural gerrymandering", we wish to highlight a potentially important distinction between \textit{neutral} plans and \textit{fair} plans -- each of which reflects a different approach to defining the baseline for what constitutes gerrymandering. Neutral plans refer to those that are drawn entirely with respect to traditional good government criteria, with no attention paid to partisan considerations. One way to define partisan gerrymandering is with respect to a baseline defined by the set of feasible neutral plans -- and this was how one of the plaintiff’s experts in the Pennsylvania case, Professor Jowei Chen \citep[cf.][]{Chen2015_JOP} shaped his testimony in the Pennsylvania case, and in other cases in which he has testified.

@@

    In contrast, we may wish to use the term ``fair" as plans that seek to compensate for the difference in level of partisan concentration. Fair plans do not seek to achieve proportional representation. Rather, they seek to achieve partisan symmetry. A fair plan would yield a proportional result when the vote shares are evenly split 50-50, but at points beyond 50-50, such as when party R wins 70\% of the districts with only 55\% of the vote, party L can similarly win 70\% of the districts if it had received 55\% of the vote. 

@@

    But the Pennsylvania Supreme Court opted not to do that, but to instead preserve county lines and draw a good government map. We therefore identify the Court remedial plan as a ``neutral" plan.

@@

    While the Pennsylvania court opinion is limited to Pennsylvania, and thus it might have seemed of importance only in Pennsylvania, in footnote 71 of the Opinion (slip op. pp.116-117), the court took what we regard as a rather unusual step. It issued what can only be called an invitation to other state courts to use the same logic it used to invalidate partisan gerrymanders in their own state. As we noted earlier, the Court pointed out that there are twelve additional states whose constitutions contain election clauses identical to the Pennsylvania charter, requiring elections to be ``free and equal”: Arizona, Arkansas, Delaware, Illinois, Indiana, Kentucky, Oklahoma, South Dakota, Oregon, Tennessee, Washington, and Wyoming.

@@

    However, only a handful of these states are ripe for partisan gerrymandering challenges -- in that (ca. 2017) three lack unified party control of the state, two are single district states, and some already have a commission drawing plans and in others, \textit{indicia} of gerrymandering are missing.

@@

    In looking to the future, we should also note that some states that are generally regarded as among the most pernicious partisan gerrymanders, Michigan, Ohio, North Carolina, Wisconsin are not included among the twelve, nor is Maryland. On the other hand, we should also note that a ``''free and equal” elections clause is not the only avenue state courts might use to attack partisan gerrymanders in the future. As University of Kentucky College of Law Professor Joshua Douglas has pointed out, virtually every state constitution protects voting rights more explicitly than the U.S. Constitution does \citep{Douglas2014_RightToVote}. In addition to the thirteen states that require elections to be ``free and equal”, an additional thirteen have state constitutional provisions that require elections to be ``free" or ``free and open”, and these clauses could, in principle, be used in exactly the same way as the ``free and equal” clause. 

@@

    Such clauses are found in a number of the states widely regarded as having the worst congressional gerrymandering. 

@@

    Accordingly, some scholars have suggested turning to turn away from the federal judiciary altogether and instead focusing on state-level litigation \citep{Wang_et_al_2019_Labortories_UPJCL}. We believe that the \textit{LWV} decision and the further comparative analyses of alternative plans presented in the previous sections of this paper will help other state courts navigate their way to decisions that strike down some partisan gerrymanders as unconstitutional, while allowing others to remain in place on the grounds that either they are not that severe, or they are unlikely to be lasting in that they have sufficiently many competitive seats as to be responsive to realistic changes in voting patterns. Now that litigation has been successful in state courts on the grounds of egregious deviations from good government districting criteria, it seems plausible that in the next wave of redistricting, mapmakers may create covert gerrymanders like we've described in this essay. If so, there remains the issue of whether state courts have the legal tools to deal with these ``stealth gerrymander”.

@@

    According to all four measures, the joint legislative plan offered by Republican leaders is nearly as much a gerrymandered map -- in a pro-Republican direction -- as is the 2011 map. As noted above, we believe it appropriate to label this plan as a \textit{stealth gerrymander} since it creates partisan assymetries while (mostly) providing the visual appearance of a good government map (see Figure \ref{fig:maps}). However, having labeled the joint legislative plan a \textit{stealth gerrymander}, it is worth reminding readers of two important points. First, had this plan been the original plan adopted by Republicans, it is very much an open question as to whether it would have been rejected by the Pennsylvania Supreme Court. Even though, in partisan terms, it is nearly identical to the 2011 plan, it is considerably more consistent with good government criteria (see Table \ref{tab:summaries}) -- not that much worse than the Court Remedial plan -- that it would have required detailed political election analyses like those we have done here to demonstrate its \textit{stealth gerrymander} features, plus a court willing to rule on political grounds, rather than good government grounds that it was unconstitutional. Because the joint legislative plan was not passed by the State of Pennsylvania's legislature, the Court felt no need for particular deference to it (or to the Governor's plan, for that matter). Thus, the Court was not bound by the normal supposition with respect to districting that a plan authorized by the state need not be the ``best possible", but only unconstitutional. That ``deference" to legislative judgments and the criteria they reflect, might well have tipped the balance toward acceptance of the Republican joint legislative plan were it offered as having been fully sanctioned by the State of Pennsylvania (the duly elected legislature and Governor). 

@@

% =====================================================================

@@

	Thus, we have some optimism that state courts can provide a venue to check partisan gerrymandering, even when it comes in the form of a \textit{stealth gerrymander}. But we have also placed a number of caveats on any claim that state courts can compensate fully for any continued failure of the federal courts to act. Citizen-based reforms such as referendum and initiatives may also help with reducing egregious gerrymandering in states where the passage of such measures is feasible. Ultimately, there is no substitute for a federal standard that guarantees each citizen in the United States \textit{equal protection under the law}. Absent the United States Supreme Court setting this standard, state courts offer one promising pathway towards ending undemocratic partisan gerrymandering.

@@

% %        ================================================================= 
% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % ------------------------------------------------------------------------
% 		\footnote{A long literature in political science has evaluated and confirmed the presence and increased incumbency advantage \citep{Abramowitz1991,Levitt1997,Ansolabehere2002b,Gelman_Huang_2008}. Recent scholarship, however, has shown a decrease in the advantage towards zero \citep{Jacobson2015_JOP}.}
% % ------------------------------------------------------------------------ 
% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% %        =================================================================

%     A well-financed incumbent tends to suppress quality competition \citep{Jacobson2015_JOP}, vulnerable incumbents who can sense the political winds opt against running for re-election \citep{Jacobson_Kernell_1981,Ansolabehere_Snyder_2004_LSQ}, and electoral tides and varying turnout can distort the state-wide average. Midterm elections often have vastly different electorates than presidential elections \citep{Abramowitz_et_al_1986_AJPS}, while the midterm pattern of ``surge and decline" (what goes up must come down principle) reduces the sitting Presidents' party's vote share \citep{CampbellA1960}. 
%  \par
% 	What we are after is a measure of the ``normal vote''. The normal vote is the underlying partisanship of a district \citep{Converse1966}. It is sort of a mystical number since it is uncovered and unobserved, but powerful theoretically. Map builders implicitly assume a normal vote when constructing a district plan. We believe that state-wide elections offer the most useful approximation of the normal vote.
% %        ================================================================= 
% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % ------------------------------------------------------------------------
% 		\footnote{We particularly believe this is true when the votes between elections are highly correlated, and the district-level variation is low, as it was in Pennsylvania from 2012 to 2016. The more stable the electorate is, the more confidence we have that state-wide elections are revealing the normal vote \citep{Ansolabehere_et_al_2000_AJPS}.}
% % ------------------------------------------------------------------------ 
% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% %        =================================================================
% 	 All districts vote on the same candidates; hence outcome affects all voters equally. The effects of incumbency can distort the vote in congressional races in a way that doesn't resemble the actual partisan preferences at the district-level. State-wide elections eliminate the need to impute or otherwise deal with uncontested districts. Taking an average of the state-wide elections provides additional reason to believe it's capturing a normal vote due to a regression to the mean effects. Justices in \textit{Vieth} offered skepticism in using elections that have contrasting results; we do not share this view, instead seeing it as an important source of variation that helps to recognize the underlying partisanship in which \textit{should} be used in measuring bias in districting plans. Additionally, averaging over multiple races and years helps to reduce the bias imposed by candidate effects.
% \par
% 	We construct the five-election composite of state-wide elections by aggregating voting district level (precincts) data to the congressional district level.

@@

% %        ================================================================= 
% %        -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% % ------------------------------------------------------------------------	
% 	We calculate this average only for districts contested in 2016, since uncontested results have significant deviations. Including the uncontested districts increases this mean difference to just 3.93\%. We believe our composite measure offers better representation of the true partisanship in these districts. Individually, only the Auditor's race correctly predicts all the districts correctly, but by either summing or by taking the average of the five races, both methods correctly predict each district's outcome.
% % ------------------------------------------------------------------------ 
% %        -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% %        =================================================================

%     We imputed the values 25\% and 75\% for uncontested or essentially uncontested races. Rather than taking the projected outcome we arrive at from this method as a single value, we looked at the sensitivity of this projection to the empirical range of inter-election swing, simulating outcomes 1,000 times. To arrive at our estimate for inter-election variation that can't be attributed to the normal vote or incumbency advantage, we estimate the equation using congressional elections in Pennsylvania in the 2004-16 time period:
%         \begin{align*}
%         \mathrm{Vote}_{ty} = \alpha + \beta_{1} \mathrm{Vote}_{yt-1} + \beta_{2} \mathrm{Incumbency}_{yt} + \beta_{4} \mathrm{Uncontestedness}_y{t} + \epsilon
%         \end{align*}
%     We are interested in $\epsilon$, which is the unexplained variance, which includes inter-election swing. We run this equation separately for each of the non-redistricting years and then average the coefficients. We estimate incumbency advantage to average 4.27\% for this time period, while inter-election variance is 4.46\%. When simulating elections, we use the composite as the normal vote, but add in for each district a random draw from a normal distribution centered at the district vote with a standard deviation of $\epsilon$. We can then calculate statistics for each of these simulated elections to find a distribution for each measure of gerrymandering.
